---
title: "Introductory Univariate Statistics in Psychology Performed as General Linear Models (GLM) in R"
subtitle: "A terse comparison of typical introductory univariate statistics in psychology performed as GLMs in R"
date: "Date Posted"
output:
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psych)
```

*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>

<BR><BR>

Note that the test statistic and its respective *p*-value are identical using either method (i.e., the typical univariate introductory statistics for psychology vs. GLM). Also, note that the GLM provides the same information as the ANOVA with the addition of 1 degree of freedom (*df*) tests. Thus, to take advantage of the 1 *df* test with categorical predictors (e.g., in the ANOVA), meaningful *a priori* contrasts are preferred over pre-programmed contrasts, especially in research.

<BR><BR>

## One sample *t*-test

```{r dataset_one_sample_ttest}

x <- rnorm(1:10)
dataset <- data.frame(x)
describe(dataset)

```

<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCell" style="width: 50%">

<CENTER>**Typical**</CENTER>

```{r}

t.test(x)

```

</div>
<div class="divTableCell" style="width: 50%">

<CENTER>**GLM**</CENTER>

```{r}

model <- lm(x ~ 1)
summary(model)
confint(model)

```

</div>
</div>

</div>
</div>

<BR><BR><BR>

## Dependent samples *t*-test

```{r dataset_dependent_samples_ttest}

x1 <- rnorm(1:10)
x2 <- rnorm(1:10)
diff <- x2-x1
dataset <- data.frame(x1, x2, diff)
dataset
describe(dataset)

```

<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCell" style="width: 50%">

<CENTER>**Typical**</CENTER>

```{r}

t.test(diff)

```

</div>
<div class="divTableCell" style="width: 50%">

<CENTER>**GLM**</CENTER>

```{r}

model <- lm(diff ~ 1)
summary(model)
confint(model)

```

</div>
</div>

</div>
</div>

<BR><BR><BR>

## Independent samples *t*-test

```{r dataset_independent_samples_ttest}

y <- rnorm(1:10)
x <- c("group1","group1","group1","group1","group1","group2","group2","group2","group2","group2")
dataset <- data.frame(y, x)
dataset
describeBy(dataset$y, group=dataset$x)

```

<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCell" style="width: 50%">

<CENTER>**Typical**</CENTER>

```{r}

t.test(y ~ x, var.equal=TRUE)

```

</div>
<div class="divTableCell" style="width: 50%">

<CENTER>**GLM**</CENTER>

```{r}

model <- lm(y ~ x)
summary(model)
confint(model)

```

</div>
</div>

</div>
</div>

<BR><BR><BR>

## One-way ANOVA

```{r dataset_one_way_anova}

y <- rnorm(1:15)
x <- c("group1","group1","group1","group1","group1","group2","group2","group2","group2","group2","group3","group3","group3","group3","group3")
dataset <- data.frame(y, x)
dataset
describeBy(dataset$y, group = x)

```

<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCell" style="width: 50%">

<CENTER>**Typical**</CENTER>

```{r}

model <- aov(y ~ x)
anova(model)

```

</div>
<div class="divTableCell" style="width: 50%">

<CENTER>**GLM**</CENTER>

```{r}

model <- lm(y ~ x)
anova(model)
summary(model)
confint(model)

```

</div>
</div>

</div>
</div>

<BR><BR><BR>

## Factorial ANOVA

```{r dataset_factorial_anova}

y <- rnorm(1:15)
x1 <- c("group1","group1","group1","group1","group1","group2","group2","group2","group2","group2","group3","group3","group3","group3","group3")
x2 <- c("group1","group2","group3","group1","group2","group3","group1","group2","group3","group1","group2","group3","group1","group2","group3")
dataset <- data.frame(y, x1, x2)
dataset
describeBy(dataset$y, group = list(dataset$x1, dataset$x2))

```

<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCell" style="width: 50%">

<CENTER>**Typical**</CENTER>

```{r}

model <- aov(y ~ x1*x2)
anova(model)

```

</div>
<div class="divTableCell" style="width: 50%">

<CENTER>**GLM**</CENTER>

```{r}

model <- lm(y ~ x1*x2)
anova(model)
summary(model)
confint(model)

```

</div>
</div>

</div>
</div>

<BR><BR><BR>

## Correlation

```{r dataset_correlation}

y <- rnorm(1:10)
x <- rnorm(1:10)
dataset <- data.frame(y,x)
dataset
describe(dataset)

```

<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCell" style="width: 50%">

<CENTER>**Typical**</CENTER>

```{r}

cor.test(x, y)

```

</div>
<div class="divTableCell" style="width: 50%">

<CENTER>**GLM**</CENTER>

```{r}

model <- lm(scale(y) ~ scale(x))
anova(model)
summary(model)
confint(model)

```

</div>
</div>

</div>
</div>