---
title: "Blog"
subtitle: "Mostly a reference"
output: html_document
---

<BR>

```{r blog_figure, echo = F, warning = F, message = F, fig.width = 3, fig.height = 2, fig.asp = T, dev='svg'}
packages <- c("readr", "glue", "tidyverse", "psych", "zoo")
xfun::pkg_attach(packages, message = F)
```

```{r, echo = F, message = F}
blogs <- read_csv("blogs.csv") %>%
  rowwise() %>%
  mutate(date = paste0(year,"-",month,"-",day)) %>%
  ungroup() %>%
  mutate(date = as.Date(date))

blogs_summary <- blogs %>%
  group_by(year,month) %>%
  summarize(frequency = n()) %>%
  rowwise() %>%
  mutate(date = paste0(year,"-",month,"-01")) %>%
  ungroup() %>%
  mutate(date = as.Date(date))

# fill in missing dates
blog_dates <- as.data.frame(seq(from = as.Date(as.yearmon(min(blogs_summary$date))), 
                 to = as.Date(as.yearmon(max(blogs_summary$date))), 
                 by="month"))
colnames(blog_dates) = "date"

blog_summary_complete <- full_join(blog_dates, blogs_summary, by = "date")
blog_summary_complete <- blog_summary_complete %>%
  rowwise() %>%
  mutate(frequency = ifelse(is.na(frequency),0,frequency)) %>%
  select(-c(year,month))

blog_mean <- describe(blog_summary_complete$frequency)["mean"]
blog_se <- describe(blog_summary_complete$frequency)["se"]

# ggplot(blog_summary_complete, aes(x = date, y = frequency)) +
#   geom_line(color = "#106a8c") +
#   geom_point(color = "#106a8c") +
#   theme_classic() +
#   scale_y_continuous(limits = c(0,max(blog_summary_complete$frequency)), breaks = c(0,1,2)) +
#   scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
#   labs(subtitle = paste("Mean Per Month =",blog_mean,"±",blog_se),
#        x = "",
#        y = "Frequency") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

<div class="divTable">
<div class="divTableBody">

<div class="divTableRow"><div class="divTableCell">
**Date**
</div><div class="divTableCell">
**Post**
</div></div>

<div class="divTableRow"><div class="divTableCell">
2019-03-07
</div><div class="divTableCell">
<div class="divBlogTitle">
<a href="https://mattkmiecik.com/post-Exploring-11-Years-of-Chicago-Blackhawks-Data-using-Principal-Components-Analysis.html" target="_blank">Exploring 11 Years of Chicago Blackhawk’s Data using Principal Components Analysis</a>
</div>
A collaboration post with <a href="https://mattkmiecik.com" target="_blank">Matt Kmiecik</a> using principal components analysis to analyze Blackhawk's data.
<BR>*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>
</div></div>

<BR>

<div class="divTableRow"><div class="divTableCell">
2019-01-30
</div><div class="divTableCell">
<div class="divBlogTitle">
[Solving Ordinary Least Squares (OLS) Regression Using Matrix Algebra](blog_solving_ols_regression_using_matrix_algebra.html)
</div>
A brief guide to solving ordianry least squares (OLS) regression using matrix algebra in R.
<BR>*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>
</div></div>

<BR>

<div class="divTableRow"><div class="divTableCell">
2018-09-01
</div><div class="divTableCell">
<div class="divBlogTitle">
<a href="https://mattkmiecik.shinyapps.io/boot-perm-dash/" target="_blank">Bootstrapping and Permutation Testing: A Shiny App (A Dashboard Update)</a>
</div>
Upate to the previous "Bootstrapping and Permutation Testing: A Shiny App" post using `shinydashboard`.
<BR>*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>
<span class="badge badge-pill badge-primary">Shiny Apps</span>
</div></div>

<BR>

<div class="divTableRow"><div class="divTableCell">
2018-08-15
</div><div class="divTableCell">
<div class="divBlogTitle">
<a href="https://epongpipat.shinyapps.io/lm_voice_dashboard/" target="_blank">Simple Linear Regression via Voice Command: A Shiny App (A Dashboard Update)</a>
</div>
Upate to the previous "Simple Linear Regression via Voice Command: A Shiny App" post using `shinydashboard`. *(Note: Link may or not work depending on if my free hours have been used up for the month, but feel free to download the code and run locally from my <a href="https://github.com/epongpipat/shiny_apps/tree/master/lm_voice_dashboard" target="_blank">GitHub <span class="fa fa-github"></span></a>.)*
<BR>*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>
<span class="badge badge-pill badge-primary">Shiny Apps</span>
</div></div>

<BR>

<div class="divTableRow"><div class="divTableCell">
2018-06-17
</div><div class="divTableCell">
<div class="divBlogTitle">
<a href="https://epongpipat.shinyapps.io/lmvoice" target="_blank">Simple Linear Regression via Voice Command: A Shiny App</a>
</div>
Change variables of a simple linear regression via voice commands. *(Note: Link may or not work depending on if my free hours have been used up for the month, but feel free to download the code and run locally from my <a href="https://github.com/epongpipat/epongpipat.github.io/tree/master/shinyApps/lmVoice" target="_blank">GitHub <span class="fa fa-github"></span></a>.)*
<BR>*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>
<span class="badge badge-pill badge-primary">Shiny Apps</span>
</div></div>

<BR>

<div class="divTableRow"><div class="divTableCell">
2018-06-07
</div><div class="divTableCell">
<div class="divBlogTitle">
<a href="https://mattkmiecik.com/post-Bootstrapping-and-Permutation-Testing-Shiny-App.html" target="_blank">Bootstrapping and Permutation Testing: A Shiny App</a>
</div>
A collaboration post with [Matt Kmiecik](https://mattkmiecik.com) about the differences between bootstrapping and permutation testing with input variables that change on the fly.
<BR>*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>
<span class="badge badge-pill badge-primary">Shiny Apps</span>
</div></div>

<BR>

<div class="divTableRow"><div class="divTableCell">
2018-03-20
</div><div class="divTableCell">
<div class="divBlogTitle">
[Calculate 24-Motion Variables](blog_Calculate_Motion_Regressors.html)
</div>
A quick guide to calculating 24-motion regressors (with visuals).
<BR>*Tags:* <span class="badge badge-pill badge-dark">fMRI</span>
<span class="badge badge-pill badge-primary">R</span>
</div></div>

<BR>

<div class="divTableRow"><div class="divTableCell" style="width: 100px;">
2018-02-11
</div><div class="divTableCell">
<div class="divBlogTitle">
<a href="https://mattkmiecik.com/post-Multilevel-Modeling-in-R-with-NHL-Power-Play-Data.html" target="_blank">Multilevel Modeling in R with NHL Power Play Data</a>
</div>
A collaboration post with [Matt Kmiecik](https://mattkmiecik.com) about multilevel modeling in R using professional hockey data from the NHL.
<BR>*Tags:* <span class="badge badge-pill badge-light">Statistics</span>
<span class="badge badge-pill badge-primary">R</span>
</div></div>

</div>
</div>