---
title: "Calculate Motion Regressors"
subtitle: "A quick guide to calculating motion regressors"
date: "2018-03-13"
output: 
  html_document:
    highlight: textmate
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Tags:* <span class="badge badge-pill badge-dark">fMRI</span>
<span class="badge badge-pill badge-primary">R</span>

<BR><BR>

## 6-Rigid Body Motion Regressors

The first six motion parameters are motion parameters that are obtained from a software of choice and include:

1. x, translation in x-direction
2. y, translation in the y-direction
3. z, translation in the z-direction
4. roll, rotation around the x-axis
5. pitch, rotation around the y-axis
6. yaw, rotation around the z-axis

```{r}

```

<BR><BR>

## 12-Motion Regressors

The next six motion parameters are the original six rigid body motion parameters squared and include:

7. x^2, translation in x-direction squared
8. y^2, translation in the y-direction squared
9. z^2, translation in the z-direction squared
10. roll^2, rotation around the x-axis squared
11. pitch^2, rotation around the y-axis squared
12. yaw^2, rotation around the z-axis squared

```{r}

```

<BR><BR>

## 18-Motion Regressors

The next six motion parameters are the first temporal derivates of the original six rigid body motion parameters. Specifically, it is the motion minus the motion from the previous time point.

13. x_td, first temporal derivative of translation in the x-direction
14. y_td, first temporal derivative of translation in the y-direction
15. z_td, first temporal derivative of translation in the z-direction
16. roll_td, first temporal derivative of rotation around the x-axis
17. pitch_td, first temporal derivative of rotation around the y-axis
18. yaw_td, first temporal derivative of rotation around the z-axis

```{r}

```

<BR><BR>

## 24-Motion Regressors

The next six motion parameters are the first temporal derivatives squared.

19. x_td_sq, first temporal derivative of translation in the x-direction squared
20. y_td_sq, first temporal derivative of translation in the y-direction squared
21. z_td_sq, first temporal derivative of translation in the z-direction squared
22. roll_td_sq, first temporal derivative of rotation around the x-axis squared
23. pitch_td_sq, first temporal derivative of rotation around the y-axis squared
24. yaw_td_sq, first temporal derivative of rotation around the z-axis squared

```{r}

```

<BR><BR>

## 30-Motion Regressors

The next six motion parameters are the second temporal derivatives of the original six rigid body motion parameters. Specifically, it is the motion minus the motion from the two previous time points.

25. x_td2, second temporal derivative of translation in the x-direction 
26. y_td2, second temporal derivative of translation in the y-direction 
27. z_td2, second temporal derivative of translation in the z-direction 
28. roll_td2, second temporal derivative of rotation around the x-axis 
29. pitch_td2, second temporal derivative of rotation around the y-axis 
30. yaw_td2, second temporal derivative of rotation around the z-axis

```{r}

```

<BR><BR>

## 36-Motion Regressors

The next six motion parameters are the second temporal derivatives squared.

31. x_td2_sq, second temporal derivative of translation in the x-direction squared
32. y_td2_sq, second temporal derivative of translation in the y-direction squared
33. z_td2_sq, second temporal derivative of translation in the z-direction squared
34. roll_td2_sq, second temporal derivative of rotation around the x-axis squared
35. pitch_td2_sq, second temporal derivative of rotation around the y-axis squared
36. yaw_td2_sq, second temporal derivative of rotation around the z-axis square

```{r}

```

<BR><BR>

## R Script
```{r}

# load 6-motion parameters
m1 <- rnorm(1:10)
m2 <- rnorm(1:10)
m3 <- rnorm(1:10)
m4 <- rnorm(1:10)
m5 <- rnorm(1:10)
m6 <- rnorm(1:10)

dataset <- data.frame(m1,m2,m3,m4,m5,m6)
colnames(dataset) <- c("x","y","z","roll","pitch","yaw")

# 12-motion parameters

dataset.sq <- dataset^2 # square 6-rigid body motion parameters
colnames(dataset.sq) <- c("x.sq","y.sq","z.sq","roll.sq","pitch.sq","yaw.sq") # rename columns

dataset.12mp <- data.frame(dataset,dataset.sq) # combine the 12-motion parameters into one dataset

# 24-motion parameters

dataset.td1 <- sapply(dataset, FUN=function(x) c(NA,diff(x))) # subtract motion from prior motion time point
colnames(dataset.td1) <- c("x.td1","y.td1","z.td1","roll.td1","pitch.td1","yaw.td1") # rename columns 

dataset.td1.sq <- dataset.td1^2 # square td1
colnames(dataset.td1.sq ) <- c("x.td1.sq ","y.td1.sq ","z.td1.sq ","roll.td1.sq ","pitch.td1.sq ","yaw.td1.sq ") # rename columns

dataset.24mp <- data.frame(dataset.12mp,dataset.td1,dataset.td1.sq) # combine the 24-motion parameters into one dataset

# 36-motion parameters

dataset.td2 <- sapply(dataset, FUN=function(x) c(NA,NA,diff(x, lag=2))) # subtract motion from the motion of two prior time points
colnames(dataset.td2) <- c("x.td2","y.td2","z.td2","roll.td2","pitch.td2","yaw.td2") # rename columns

dataset.td2.sq <- dataset.td2^2 # square td2
colnames(dataset.td2.sq ) <- c("x.td2.sq ","y.td2.sq ","z.td2.sq ","roll.td2.sq ","pitch.td2.sq ","yaw.td2.sq ") # rename columns 

dataset.36mp <- data.frame(dataset.24mp,dataset.td2,dataset.td2.sq) # combine the 36-motion parameters into one dataset

```

<BR><BR>

## References
[1]: 