---
title: "Curriculum Vitae"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
packages <- c("tidyverse", "knitr", "kableExtra", "glue", "psych", "scales", "eepR", "icon")
xfun::pkg_attach(packages, install = T, message = F)
```

*Updated on February 25, 2020*
<BR><BR><BR>

## Education
<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_utd.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>Ph.D. Student, Cognition and Neuroscience</B> (2017 - Present)
<BR> The University of Texas at Dallas
<BR> Advisors: Drs. Karen M. Rodrigue & Kristen M. Kennedy
</div>
</div>

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_sdsu.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>M.A., Psychology</B> (2016)
<BR> San Diego State University
<BR> Advisor: Dr. Claire Murphy
<BR> Thesis: Altered Functional Connectivity and Eating Disinhibition in Individuals with Metabolic Syndrome
</div>
</div>

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_csun.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>B.A., Psychology</B> (2012)
<BR> California State University, Northridge
<BR> Advisor: Dr. Jose P. Abara
</div>
</div>

</div>
</div>

<BR><BR>

## Research Experience
<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_utd.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>Graduate Research Assistant</B> (2017 - Present)
<BR>Cognitive Neuroscience of Aging Laboratory
<BR>Center for Vital Longevity
<BR>School of Behavioral and Brain Sciences
<BR>The University of Texas at Dallas
</div>
</div>

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_sdsu.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>Graduate Research Assistant</B> (2014 - 2016)
<BR>Lifespan Human Senses Laboratory
<BR>Department of Psychology
<BR>San Diego State University
</div>
</div>

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_csun.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>Undergraduate Research Assistant</B> (2011 - 2013)
<BR>Neuroscience Laboratory
<BR>Department of Psychology
<BR>California State University, Northridge
</div>
</div>

</div>
</div>

<BR><BR>

## Teaching Experience
<div class="divTable">
<div class="divTableBody">

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_utd.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>Graduate Teaching Assistant</B> (2017 - 2019)
<BR>The University of Texas at Dallas
<BR>1. Statistics for Psychology, ACN6312/HCS6312 (Spring 2019)
<BR><div class="divIndent">Worked with Dr. Nancy Juhn</div>
<BR>2. Statistics for Psychology, PSY2317 (Fall 2017, Spring 2018, Fall 2018)
<BR><div class="divIndent">Worked with Dr. Nancy Juhn</div>
</div>
</div>

<div class="divTableRow">
<div class="divTableCellLogo">
<img src="images/logo_sdsu.svg" width="100">
</div>
<div class="divTableCellInfo"> 
<B>Graduate Teaching Assistant</B> (2014 - 2016)
<BR>San Diego State University
<BR>1. Advanced Statistics in Psychology, PSY670A/B (Fall 2015, Spring 2016)
<BR><div class="divIndent">Worked with Dr. Melody Sadler</div>
<BR>2. Statistical Methods in Psychology Laboratory, PSY280L (Fall 2014, Spring 2015)
<BR><div class="divIndent">Worked with Dr. Melody Sadler, Dr. Amy Spilkin, Dr. Dale Glaser</div>
</div>
</div>

</div>
</div>

<BR><BR>

## Publications

<div class="divReferences">
McIntosh, E. C., Jacobson, A., Kemmotsu, N., **Pongpipat, E. E.**, Green, E., Haase, L., & Murphy, C. (2016). Does medial temporal lobe thickness mediate the association between risk factor burden and memory performance in middle-aged or older adults with metabolic syndrome? *Neuroscience Letters, 636*, 1-9. http://doi.org/10.1016/j.neulet.2016.10.010
</div>

<BR><BR>

## Poster Presentations

<div class="divReferences">
```{r posters, warning = F, message = F, fig.width = 5, fig.height = 2, fig.asp = T, dev='svg'}
posters <- read_csv("posters.csv")

posters$Authors <- gsub("Pongpipat, E. E.", "<b>Pongpipat, E. E.</b>", posters$Authors)

posters <- posters %>%
  rowwise() %>%
  mutate(Date = ifelse(Status != "Complete", paste(Year, Month, tolower(Status), sep = ", "), paste(Year, Month, sep = ", ")))

posters <- posters %>%
  mutate(html = paste0(Authors, " (", Date, "). <i>", Title, "</i>. Poster session presented at the ", Conference, ", ", Location, ifelse(is.na(Note), "", paste0(". (", Note, ")."))))

posters_html <- paste(posters$html, collapse = "\n\n")

poster_summary <- posters %>%
  select(Year) %>%
  group_by(Year) %>%
  summarize(Frequency = n())

poster_summary <- rbind(poster_summary, c(2017, 0))

poster_summary <- poster_summary %>%
  arrange(Year)

poster_mean_per_year <- round(describe(poster_summary$Frequency)[["mean"]], 2)
poster_se_per_year <- round(describe(poster_summary$Frequency)[["se"]], 2)
poster_total <- sum(poster_summary$Frequency)

ggplot(poster_summary, aes(x = as.Date(paste0(poster_summary$Year,"-01-01"), "%Y-%m-%d"), y = Frequency)) +
  geom_line(color = "#106a8c") +
  geom_point(color = "#106a8c") +
  theme_classic() +
  scale_x_date(labels = date_format("%Y"), date_breaks= "1 year") +
  labs(subtitle = paste0("Total: ", poster_total, ", Mean Per Year: ",poster_mean_per_year," ± ",poster_se_per_year),
       x = "Year") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

`r posters_html`

</div>

<BR><BR>

## Talks and Workshops

```{r talks, results = 'asis'}
talks <- tribble(~Date, ~Description, ~Location,
                 "2019-11", "Brianhack Dallas 2019 <br><a href='https://github.com/Brainhack-Dallas/mini-brainhack-utd' target='blank'><i class='fab fa-github'></i></a> &#8201 <a href='https://brainhack-dallas.github.io/mini-brainhack-utd/' target='_blank'><i class='fas fa-link'></i></a>", "UT Dallas",
                 "2019-10", "Talk on 'Functional connectivity of the fronto-parietal and default mode networks during the n-back across the adult lifespan' given at the Developmental, Cognitive and Social/Personality Brownbag Series", "UT Dallas",
                 
                 "2019-09", "Assisted in an introduction to Git/GitHub Workshop with Dr. Micaela Chan <br><a href='https://github.com/mychan24/git_github_workshop' target='_blank'><i class='fab fa-github'></i></a> &#8201 <a href='https://gitbookdown.site' target='_blank'><i class='fas fa-book'></i></a>", "Center for Vital Longevity<br>UT Dallas",
                 
                 "2019-04", "Guest lectured on using R for descriptive and inferential statistics for Dr. Juhn's Research Methods 2 Course", "UT Dallas")

kable(talks, escape = F) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, width = "80px") %>%
  column_spec(3, width = "175px")
```

<BR><BR>

## Service to the Profession

Co-reviewer for NeuroImage with Dr. Kennedy

<BR><BR>

## Honors and Awards
```{r honors_awards}
honors_awards <- tibble("2017 (Fall)", "CSU Northridge", "Dean's List")

# test <- cbind(2016, "WPA Robert L. Solso Award")
colnames(honors_awards) <- c("Year", "Institution", "Award")

honors_awards <- honors_awards %>%
  add_row(., Year = "2010 (September)", Institution = "Cope Health Solutions", Award = "Clinical Care Extender Intern of the Rotation Award", .before = 1) %>%
  add_row(., Year = "2010 (Spring)", Institution = "CSU Northridge", Award = "Dean's List", .before = 1) %>%
  add_row(., Year = "2011 (Fall)", Institution = "CSU Northridge", Award = "Dean's List", .before = 1) %>%
  add_row(., Year = "2011 (Spring)", Institution = "CSU Northridge", Award = "Dean's List", .before = 1) %>%
  add_row(., Year = "2012 (Fall)", Institution = "CSU Northridge", Award = "Dean's List", .before = 1) %>%
  add_row(., Year = "2014 (January)", Institution = "Huntington Learning Center", Award = "Teacher of the Month", .before = 1) %>%
  add_row(., Year = "2014", Institution = "Western Psychological Association", Award = "Student Scholarship Award", .before = 1) %>%
  add_row(., Year = "2014 (June)", Institution = "Huntington Learning Center", Award = "Teacher of the Month", .before = 1) %>%
  add_row(., Year = "2015", Institution = "Western Psychological Association", Award = "Robert L. Solso Award", .before = 1) %>%
  add_row(., Year = "2015", Institution = "San Diego State University", Award = "Dr. Edward Geldreich Scholarship", .before = 1) %>%
  add_row(., Year = "2016", Institution = "Western Psychological Association", Award = "Robert L. Solso Award", .before = 1) %>%
  add_row(., Year = "2016", Institution = "San Diego State University", Award = "Honorable Mention for Outstanding Service as a Teaching Assistant", .before = 1) %>%
  add_row(., Year = "2018", Institution = "The University of Texas at Dallas", Award = "Student Travel Award", .before = 1) %>%
  add_row(., Year = "2020", Institution = "The University of Texas at Dallas", Award = "Student Travel Award", .before = 1)
# add_row(., Year = "", Institution = "", Award = "", .before = 1) %>%


# Prints table
kable(honors_awards, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
  column_spec(1, width = "150px")
```

<BR><BR>

## Professional Society Membership
```{r professional_society_membership}

professional_society_membership <- tibble("2012 - 2016", "Western Psychological Association")

# test <- cbind(2016, "WPA Robert L. Solso Award")
colnames(professional_society_membership) <- c("Year", "Institution")

professional_society_membership <- professional_society_membership %>%
  add_row(., Year = "2015 - 2017", Institution = "The Obesity Society", .before = 1) %>%
  add_row(., Year = "2012 - Present", Institution = "Society for Neuroscience", .before = 1) %>%
  add_row(., Year = "2016 - Present", Institution = "Phi Kappa Phi Honor Society", .before = 1)
# add_row(., Year = "", Institution = "", .before = 1) %>%

# Prints table
kable(professional_society_membership, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive")) %>%
  column_spec(1, width = "150px")
```

<BR><BR>